Fitness traits of deoxynivalenol and nivalenol-producing Fusarium graminearum species complex strains from wheat

### 2017-12-18 code.Rmd
Several changes including .doc as output

% Code

@@

The data com from a series of different experiments. There is one single `xlsx` file with each experiment in a single sheet. Follows the commented code used for these analyses separated per experiment.

## Load packages

@@

Let's first import the data and have a look at the raw data.

@@

@@

The data were prepared in the wide format where each score for percent perithecia coverage (0 -3) is in a separate column. Let's transform this dataframe from the wide to the long format using `gather` function of `dplyr`. We need to create a column for the variable name and another for the score.

@@

Let's visualize the frequency of the scores for each species-genotype separated for each substrate. This is a nice plot to show in the paper!

@@

Since we have one species which has two genotypes, it may be important, besides testing the effect of species or genotype in the model, to create a species*genotype interaction variable and fit a mixed model to test the effect of this variable on the perithecial production index (ppi). This way the two genotypes within one species can be compared between them and with the others.

@@

Now we fit the model using `lmer` function. We treat `Isolate` as a random effects.

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
glmm_per <- lmer(
  ppi ~ species_genotype * substrate + (1 | Isolate),
  data = perithecia, REML = FALSE
)
```

Let's check the single and interaction effects.

```{r message=FALSE, warning=FALSE}
Anova(glmm_per)
```

@@

```{r}
plot(glmm_per, type = c("p", "smooth"))
qqmath(glmm_per, id = 0.05)
```

@@

Not too bad. Let's now compare the means of treatments and create a dataframe which will be use to further create a plot with the estimated means and confidence interval. We used the `emmeans` package which is an update for the old `lsmeans` package. The syntax is the same as before.

```{r message=FALSE, warning=FALSE}
medias <- emmeans(glmm_per, ~ species_genotype * substrate)
med <- cld(medias, Letters = LETTERS, alpha = .05)

@@

This new `med` dataframe does not contain the two original variables we want to use to plot the data. Let's separate the `species_genotype` variable and create `genotype` and `species`.

@@

We now produce the plot with point estimate for the means and error bars for the 95%CI of the `ppi` for each `genotype`. We add `species` to the legend when using `color` argument.

@@

There is one `csv` file with data from a single experiment. We will use `read_csv` to impor the data.

@@

We will now produce the plot with the means of mycelial growth rate for each species-genotype and temperature. This will be the final plot for publication.

@@

We now create different datasets for each temperature and test the effect of `species_genotype` within each temperature.

@@

## Sporoulation & Germination

In this experiment, the production of macroconidia and further germination of 20 randomly selected spores of the different strains of the species-genotype combination, were evaluated in two runs of the experiment.

@@

Anova model

```{r message=FALSE, warning=FALSE}
lmer_spor <- lmer(spores_ml ~ species_genotype + (1 | species_genotype), data = spor, REML = FALSE)
Anova(lmer_spor)
plot(lmer_spor, type = c("p", "smooth"))
qqmath(lmer_spor, id = 0.05)
```

@@

```{r}
medias <- emmeans(lmer_spor, ~species_genotype)
med <- cld(medias, Letters = LETTERS, alpha = .05)
med <- data.frame(med)
med
```

@@

@@

## Pathogenicity tests

@@

```{r message=FALSE, warning=FALSE}
pathogen <- read_csv(here("data", "pathogenicity.csv")) %>%
  filter(isolate != "test")
```

@@

Let's prepare several summary statistics for the disease variables per cultivar. We will calculate the mean and standard deviation of proportion of diseased spikelets.

@@

pathogen_194 <- pathogen %>%
  filter(cultivar == "BRS 194") %>%
  select(1:7, 9, 11, 12, 13) %>%
  gather(position, count, 8:9) %>%
  # filter(exp == 2) %>%
  group_by(cultivar, dai, genotype, species) %>%
  summarize(
    mean_sev = mean(sev),
    sd_sev = sd(sev)
  )

@@

pathogen_Gua <- pathogen %>%
  filter(cultivar == "BRS Guamirim") %>%
  select(1:7, 9, 11:13) %>%
  gather(position, count, 8:9) %>%
  # filter(exp == 2) %>%
  group_by(cultivar, dai, genotype, species, position) %>%
  summarize(
    mean_sev = mean(sev),
    sd_sev = sd(sev)
  )
```

@@

Let's calculate the area under the curve for each isolate. For this, we will use the `audpc` function of the `agricolae` package. First, we need to group several variables up to `dai` (days after inoculation). To make our life easier, lets use the `do` in conjuction with the `tidy` functions of the `dplyr` and `broom` package, respectively. This will calculate the audpce for each spike.

@@

@@

```{r}
audpc_194 %>%
  group_by(species, genotype) %>%
  summarize(
    mean_audpc = mean(audpc),
    sd_audpc = sd(audpc)
  ) %>%
  ggplot(aes(species, mean_audpc, color = genotype)) +
  coord_flip() +
  theme_few() +
  geom_point(position = position_dodge(width = 0.4)) +
  geom_errorbar(aes(min = mean_audpc - sd_audpc, max = mean_audpc + sd_audpc), position = position_dodge(width = 0.4), width = 0.2)
```

```{r}
audpc_gua %>%
  group_by(species, genotype) %>%
  summarize(
    mean_audpc = mean(audpc),
    sd_audpc = sd(audpc)
  ) %>%
  ggplot(aes(species, mean_audpc, color = genotype)) +
  coord_flip() +
  theme_few() +
  geom_point(position = position_dodge(width = 0.4)) +
  geom_errorbar(aes(min = mean_audpc - sd_audpc, max = mean_audpc + sd_audpc), position = position_dodge(width = 0.4), width = 0.2)
```

@@

```{r}

mix_gua <- lmer(
  audpc ~ genotype +
    (1 | exp / spike),
  data = audpc_gua, REML = FALSE
)
```

@@

plot(mix_gua, type = c("p", "smooth"), col = "black")
qqmath(mix_gua, id = 0.05, col = "black")
```

```{r}
Anova(mix_gua)

means <- emmeans(mix_gua, ~ genotype)
cld(means)
```

```{r}
mix_194 <- lmer(log(audpc) ~ genotype + (1 | exp / spike), data = audpc_194, REML = FALSE)
```

```{r}

plot(mix_194, type = c("p", "smooth"), col = "black")
qqmath(mix_194, id = 0.05, col = "black")
```

@@

```{r}
Anova(mix_194)
means <- emmeans(mix_194, ~ genotype)
cld(means)
```

@@

@@

```{r message=FALSE, warning=FALSE}
ec50 <- read_csv(here("data", "ec50.csv"))
```

@@

```{r}
ec502 <- ec50 %>%
  group_by(isolate, species, genotype) %>%
  summarize(
    mean_ec50 = mean(ec50),
    sd_ec50 = sd(ec50)
  )
```

@@

Let's have a look at the mean EC50 for the isolates

```{r}
ec502 %>%
  ggplot(aes(genotype, mean_ec50, color = species)) +
  geom_jitter(width = 0.1) +
  # geom_errorbar(aes(min = mean_ec50 - sd_ec50, max = mean_ec50 + sd_ec50), width=0.2, position = position_dodge(width = 0.2))+
  theme_few() +
  theme(legend.position = "right", legend.text = element_text(size = 7, face = "italic")) +
  ylim(0, 1.8) +
  scale_color_viridis(discrete = TRUE) +
  labs(y = (expression(paste("EC"[50], " (", mu, "g/ml)"))), x = "Trichothecene genotype", color = "Species") +
  ggsave(here("figs", "ec50-new.png"), width = 4.5, height = 3)
```

>>>>

% Analysis code

@@

The data were obtained from a series of randomized replicated experiments. There is one single `csv` file that contains data from each experiment. All analyses are presented below and separated for each experiment. I follow the steps (not in that exactly order) ilustrated in the R for Data Science book (http://r4ds.had.co.nz/) (Grolemund and Wickham 2017). These are: data import and tidy, explore (transform, visualize and model) and communicate.

These are the packages used.

@@

Let's have a look at the raw data.

@@

The main response variable is ppi (perithecia production index), which was already calculated (it is an index based on the frequency of the scores). However, the original scores are included and consist of the number of grains on each score (0 to 3).

@@

The frequecy of each score is shown in separate columns. To plot and further work with the data, we will need to format the data from the wide to the long format, where each frequency of the scores need to be in a single colum. We will use the `gather` function of `dplyr`. For this, we need to inform the name of columns for the score and the frequency.

@@

We now can visualize the frequency of the scores for each species-genotype separated (facet) for each substrate. This is a nice plot to have in the paper!

@@

Since we have one species with two genotypes, it may be important, besides testing the effect of species or genotype in the model, to create a species_genotype variable. We will then fit a mixed model to test the effect species_genotype on the perithecial production index (ppi). In this way the two genotypes within one species can be compared between them and with the others. We use the `unite` function.

@@

Now we fit the mixed model using `lmer` function. We treat `Isolate` as a random effects.

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
library(lme4)
lmm_per <- lmer(
  ppi ~ species_genotype * substrate + (1 | Isolate),
  data = perithecia, REML = FALSE
)
```

Let's check the single and interaction effects using `Anova` function of the `car` package.

```{r message=FALSE, warning=FALSE}
library(car)
Anova(lmm_per)
```

@@

```{r}
plot(lmm_per, type = c("p", "smooth"))
qqmath(lmm_per, id = 0.05)
```

@@

Not too bad. Let's now compare the means of treatments and create a dataframe which will be use to further create a plot with the estimated means and confidence interval. We will use the `emmeans` package which is an update for the old `lsmeans` package. The syntax is the same as before.

```{r message=FALSE, warning=FALSE}
library(emmeans)
medias <- emmeans(lmm_per, ~ species_genotype * substrate)
med <- cld(medias, Letters = LETTERS, alpha = .05)

@@

This new `med` dataframe does not contain the two original variables we want to use to plot the data. Let's separate the `species_genotype` variable and create `genotype` and `species`. We use the `separate` function.

@@

We now can plot the point estimates for the means and respective 95%CI of the `ppi` for each `genotype`. We add `species` legends when using `color` argument so we can have five species_genotype.

@@

The main variable is mycelia growth rate, or the ratio of the growth at the fifth day and number of days. Besides the species and genotype factors, there are two levels of temperatures being tested.

@@

We can plot means of mycelial growth rate for each species-genotype and temperature as facetting variable. This plot will be included in the publication.

@@

The interaction was significant. We now create different datasets for each temperature and test the effect of `species_genotype` within each temperature.

@@

## Sporulation & Germination

In this experiment, the production of macroconidia and further germination of 20 randomly selected spores of the different strains of the species-genotype combination, was evaluated in two runs of the experiment.

@@

Mixed model for testing the effect of genotype.

```{r message=FALSE, warning=FALSE}
lmer_spor <- lmer(spores_ml ~ genotype + (1 | genotype / isolate), data = spor, REML = FALSE)

```

```{r}
plot(lmer_spor, type = c("p", "smooth"), col = "black")
qqmath(lmer_spor, id = 0.05, col = "black")
```

Model fitted well the data. Let's proceed.

```{r}
Anova(lmer_spor)
```

@@

```{r}
medias <- emmeans(lmer_spor, ~genotype)
med <- cld(medias, Letters = LETTERS, alpha = .05)
med <- data.frame(med)
med
```

@@

We will produce a combo figure for the three variables.

@@

## Pathogenicity

The pathogenicity was assessed based on the progress of the symptoms from the central (inoculated) spikelet. The data was transformed to percent severity of the spike (proportion of diseased spikelets). There is one obsevation of severity for each day after inoculation (dai) and individual spike.

@@

```{r message=FALSE, warning=FALSE}
pathogen <- read_csv(here("data", "pathogenicity.csv")) %>%
  filter(isolate != "test") %>% 
  select (-c(8:12))

pathogen
```

@@

Since we have spikes as replicates, we will calculate the mean and standard deviation of severity for plotting purposes.

@@

pathogen_194 <- pathogen %>%
  filter(cultivar == "BRS 194") %>%
  group_by(cultivar, dai, genotype, species) %>%
  summarize(
    mean_sev = mean(sev),
    sd_sev = sd(sev)
  )

@@

pathogen_Gua <- pathogen %>%
  filter(cultivar == "BRS Guamirim") %>%
   # filter(exp == 2) %>%
  group_by(cultivar, dai, genotype, species) %>%
  summarize(
    mean_sev = mean(sev),
    sd_sev = sd(sev)
  )
```

@@

Let's calculate the area under the curve for each isolate. For this, we will use the `audpc` function of the `agricolae` package. First, we need to group several variables up to `dai` (days after inoculation). To make our life easier, lets use the `do` in conjuction with the `tidy` functions of the `dplyr` and `broom` package, respectively. This will calculate the audpc for each spike.

@@

Now we can plot the means and standard deviation for each isolate.

```{r}
audpc_194 %>%
  group_by(isolate, species, genotype) %>%
  summarize(
    mean_audpc = mean(audpc),
    sd_audpc = sd(audpc)
  ) %>%
  ggplot(aes(reorder(isolate, mean_audpc), mean_audpc, color = genotype, shape = species)) +
  coord_flip() +
  theme_few() +
  geom_point(position = position_dodge(width = 0.4)) +
  geom_errorbar(aes(min = mean_audpc - sd_audpc, max = mean_audpc + sd_audpc), position = position_dodge(width = 0.4), width = 0.2)
```

@@

Now we can plot the means and standard deviation for each isolate.

```{r}
audpc_gua %>%
  group_by(isolate, species, genotype) %>%
  summarize(
    mean_audpc = mean(audpc),
    sd_audpc = sd(audpc)
  ) %>%
  ggplot(aes(reorder(isolate, mean_audpc), mean_audpc, color = genotype, shape = species)) +
  coord_flip() +
  theme_few() +
  geom_point(position = position_dodge(width = 0.4)) +
  geom_errorbar(aes(min = mean_audpc - sd_audpc, max = mean_audpc + sd_audpc), position = position_dodge(width = 0.4), width = 0.2)
```

### Mixed model

We fit a separate model for each cultivar because the experiments were conducted at different times. We fit a multilevel to test the effect of specis_genotype as fixed effects and spikes nested within isolates as random effects. The species_genotype interaction was non-significant (not shown), and then a simpler model is used.

@@

```{r}
mix_gua <- lmer(
  audpc ~ species_genotype  +
    (1 | isolate/spike),
  data = audpc_gua, REML = FALSE)

mix_gua

```

@@

```{r message=FALSE, warning=FALSE}
plot(mix_gua, type = c("p", "smooth"), col = "black")
qqmath(mix_gua, id = 0.05, col = "black")
```

The model seems OK. Let's check the significance of the species_genotype and estimated means.

```{r}
Anova(mix_gua)
means <- emmeans(mix_gua, ~ species_genotype)
cld(means)
```

We can see that the variation among isolates was too large and so no difference could be detected among the five species_genotype in the moderate resistant cultivar.

@@

We now apply the same model for the susceptible cultivar.

```{r}
mix_194 <- lmer(audpc ~ species_genotype + 
                  (1 | isolate/spike), 
                data = audpc_194)
mix_194
```

Evaluate the model.

```{r}
plot(mix_194, type = c("p", "smooth"), col = "black")
qqmath(mix_194, id = 0.05, col = "black")
```

Note: I used both the original and the log-transformed `audpc` and the results were the same. Hence, I kept the original data.

```{r}
Anova(mix_194)
means <- emmeans(mix_194, ~ species_genotype)
cld(means)
```

Again, the effect of species_genotype was not significant due to large variation among isolates.

@@

The response variable is the EC50 estimated in two experiments (true replicates) for each selected isolate from each species_genotype.

@@

```{r message=FALSE, warning=FALSE}
ec50 <- read_csv(here("data", "ec50.csv"))
ec50
```

@@

Let's obtain the mean and standard deviation of ec50 for each strain.

```{r message=FALSE, warning=FALSE}
ec502 <- ec50 %>%
   unite(species_genotype, species, genotype, sep = "_", remove = F) %>% 
  group_by(isolate, species, genotype, species_genotype) %>%
  summarize(
    mean_ec50 = mean(ec50),
    sd_ec50 = sd(ec50)
  )
ec502
```

@@

Let's have a look at the mean EC50 for each isolate.

```{r message=FALSE, warning=FALSE}
ec502 %>%
  ggplot(aes(reorder(isolate, mean_ec50), mean_ec50, shape = species, color = genotype)) +
  geom_point() +
  geom_errorbar(aes(min = mean_ec50 - sd_ec50, max = mean_ec50 + sd_ec50), width = 0.2)+
  theme_few() +
  coord_flip()+
  theme(legend.position = "right", legend.text = element_text(size = 7, face = "italic")) +
  ylim(0, 1.8) +
  scale_color_viridis(discrete = TRUE) +
  labs(y = (expression(paste("EC"[50], " (", mu, "g/ml)"))), x = "FGSC strain", shape = "Species", color = "Genotype" ) +
  ggsave(here("figs", "ec50-new.png"), width = 4.5, height = 3)
```

### 2017-12-18 index.Rmd
Several changes including .doc as output

<img src = "https://www.apsnet.org/edcenter/intropp/lessons/fungi/ascomycetes/PublishingImages/Fusariumdiscyclesm.jpg" align= right width = 250> This is a research compendium that contains and describe the data and the R codes for analyzing and visualizing these data obtained from a series of experiments conducted as an MSc research in the Del Ponte lab. The main research's objective was to assess and compare members of a species complex (FGSC), with varying toxigenic potential, that cause [Fusarium head blight or scab](https://www.apsnet.org/edcenter/intropp/lessons/fungi/ascomycetes/Pages/Fusarium.aspx), an important disease of wheat and barley worldwide. Several fitness-related traits were compared in five experiments conducted both *in vitro* and in *planta*. Check the manuscript page.

>>>>

<img src = "https://www.apsnet.org/edcenter/intropp/lessons/fungi/ascomycetes/PublishingImages/Fusariumdiscyclesm.jpg" align= right width = 270> This research compendium  contains the data and the commented R codes for analyzing and visualizing these data obtained from a series of experiments conducted as an MSc research in the Del Ponte lab.

The main objective was to assess and compare members of a species complex (FGSC), with varying toxigenic potential, which cause [Fusarium head blight (FHB) or scab](https://www.apsnet.org/edcenter/intropp/lessons/fungi/ascomycetes/Pages/Fusarium.aspx). FHB is an important disease of wheat and barley worldwide. Several fitness-related traits were compared in five experiments conducted both *in vitro* and in *planta*. Check the manuscript page.
